Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-10-18 ms/ms.tex
added results and figure for fem model

\documentclass[12pt,halfline,a4paper,]{ouparticle}

% Packages I think are necessary for basic Rmarkdown functionality
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{framed}

% Link coloring
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous Arabidopsis thaliana Incurs No Photosynthetic Advantage}
            }

%% To allow better options for figure placement
%\usepackage{float}

% Packages that are supposedly required by OUP sty file
\usepackage{amssymb, amsmath, geometry, amsfonts, verbatim, endnotes, setspace}

@@

% Macros for dealing with affiliations, footnotes, etc.
\makeatletter
\def\Newlabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}

\def\Ref#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

@@

\newcommand*\thanksbycode[1]{%
  \ifcounter{FNCT@#1}
    {\samethanks[\value{FNCT@#1}]}
    {\thanks{\Ref{FN}{#1}}\newcounter{FNCT@#1}\setcounter{FNCT@#1}{\value{footnote}}}
}

% Create labels for Addresses if the are given in Elsevier format
\Newlabel{ADR}{Colgate University}{13 Oak Drive, Hamilton, NY 13346}
\Newlabel{ADR}{University of Hawaii at Manoa}{2500 Campus Rd, Honolulu,
HI 96822}

% Create labels for Footnotes if the are given in Elsevier format
\Newlabel{FN}{1}{Equal contribution}
\Newlabel{FN}{2}{Current email address:
\href{mailto:cat@example.com}{cat@example.com}}

% Part for setting citation format package: natbib

% Part for setting citation format package: biblatex

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage{booktabs}

@@

\title{Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces
In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic
Advantage}

\author{%
%
% Code for old style authors field
%
% Add \and if both authors and author
%
%
% Code for new (elsevier) style author field
\name{Jacob L. Watts}
\address{\Ref{ADR}{Colgate University}}
%
\email{\href{mailto:jwatts@colgate.edu}{jwatts@colgate.edu}}%
\thanks{Corresponding author; Email: \href{mailto:jwatts@colgate.edu}{jwatts@colgate.edu}}%
%
%
\and
\name{Graham Dow}
\address{\Ref{ADR}{ETH}}
%
\email{\href{mailto:graham.dow@usys.ethz.ch}{graham.dow@usys.ethz.ch}}%
%
%
%
\and
\name{Thomas N. Buckley}
\address{\Ref{ADR}{University of California Davis}}
%
\email{\href{mailto:tnbuckley@ucdavis.edu}{tnbuckley@ucdavis.edu}}%
%
%
\thanksbycode{1}
%
\and
\name{Christopher D. Muir}
\address{\Ref{ADR}{University of Hawaii at Manoa}}
%
\email{\href{mailto:cdmuir@hawaii.edu}{cdmuir@hawaii.edu}}%
%
%
\thanksbycode{1}
%
%
}

\abstract{This is the abstract.

It consists of two paragraphs.}

\date{\today}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry points through which gas exchange occurs despite making up a small
percentage of the leaf area
(\protect\hyperlink{ref-lange_responses_1971}{Lange et al. 1971}).
Stomata consist of two guard cells which open and close upon changes in
turgor pressure or hormonal cues
(\protect\hyperlink{ref-mcadam_linking_2016}{McAdam and Brodribb 2016}).
The stomatal pore leads to an internal space known as the substomatal
cavity where gases contact the mesophyll. Once in the mesophyll,
CO\(_2\) diffuses throughout a network of intercellular air space (IAS)
and into mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs
within the chloroplasts (\protect\hyperlink{ref-lee_diffusion_1964}{Lee
and Gates 1964}). Stomatal conductance and transpiration are determined
by numerous environmental and anatomical parameters such as vapor
pressure deficit (VPD), irradiance, temperature, wind speed, leaf water
potential, IAS geometry, mesophyll cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
(\protect\hyperlink{ref-cowan_stomatal_1977}{Cowan and Farquhar 1977};
\protect\hyperlink{ref-buckley_optimal_2017}{Buckley, Sack, and Farquhar
2017}; \protect\hyperlink{ref-sperry_predicting_2017}{Sperry et al.
2017}). However, stomatal anatomy may be partially constrained by
physical and developments limits on phenotypic expression
(\protect\hyperlink{ref-croxdale_stomatal_2000}{Croxdale 2000};
\protect\hyperlink{ref-harrison_influence_2020}{Harrison et al. 2020};
\protect\hyperlink{ref-muir_how_2023}{Christopher D. Muir et al. 2023}).
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
(\protect\hyperlink{ref-geisler_oriented_2000}{Geisler, Nadeau, and Sack
2000}; \protect\hyperlink{ref-dow_physiological_2014}{Dow, Berry, and
Bergmann 2014}); however some species (notably in \emph{Begonia}) appear
to benefit from overlapping vapor shells caused by stomatal clustering
(\protect\hyperlink{ref-yi_gan_stomatal_2010}{Yi Gan et al. 2010};
\protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015};
\protect\hyperlink{ref-papanatsiou_stomatal_2017}{Papanatsiou, Amtmann,
and Blatt 2017}). Stomatal traits also vary adaptively in different
environments. Stomatal density positively co-varies with irradiance
during leaf development and negatively co-varies with CO\(_2\)
concentration (\protect\hyperlink{ref-gay_influence_1975}{Gay and Hurd
1975}; \protect\hyperlink{ref-schoch_dependence_1980}{Schoch, Zinsou,
and Sibi 1980}; \protect\hyperlink{ref-woodward_stomatal_1987}{Woodward
1987}; \protect\hyperlink{ref-royer_stomatal_2001}{Royer 2001}),
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
(\protect\hyperlink{ref-jordan_environmental_2015}{Jordan et al. 2015}).
Size positively co-varies with genome size
(\protect\hyperlink{ref-roddy_scaling_2020}{Roddy et al. 2020}) and
negatively co-varies with stomatal density
(\protect\hyperlink{ref-camargo_density_2011}{Camargo and Marenco
2011}). Total stomatal area (size \(\times\) density) is optimized for
operational conductance (\(gs_\text{s,op}\)) rather than maximum
conductance (\(gs_\text{s,max}\)) such that stomatal apertures are most
responsive to changes in the environment at their operational aperture
(\protect\hyperlink{ref-franks_physiological_2012}{Franks et al. 2012};
\protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Stomatal
aperture can compensate for maladaptive stomatal densities to an extent
(\protect\hyperlink{ref-bussis_stomatal_2006}{Büssis et al. 2006}), but
stomatal density and size ultimately determine a leaf's theoretical
\(gs_\text{s,max}\)
(\protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley
2016}), which is proportional to \(gs_\text{s,op}\)
(\protect\hyperlink{ref-murray_consistent_2020}{Murray et al. 2020}).
Additionally, low stomatal densities lead to irregular and insufficient
CO\(_2\) supply and reduced photosynthetic efficiency in areas far from
stomata (\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland
Pieruschka et al. 2006};
\protect\hyperlink{ref-morison_lateral_2005}{Morison et al. 2005}),
while high stomatal densities can reduce water use efficiency (WUE)
(\protect\hyperlink{ref-bussis_stomatal_2006}{Büssis et al. 2006}) and
incur excessive metabolic costs
(\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}). In
most species, stomata occur on the abaxial (usually lower) leaf surface;
but amphistomy, the occurrence of stomata on both abaxial and adaxial
leaf surfaces, is also prevalent in high light environments with
constant or intermittent access to sufficient water
(\protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary
1982}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and
Brodribb 2014}; \protect\hyperlink{ref-muir_light_2018}{Christopher D.
Muir 2018}; \protect\hyperlink{ref-drake_two_2019}{Drake et al. 2019};
\protect\hyperlink{ref-muir_is_2019}{Christopher D. Muir 2019}).
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
(\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978};
\protect\hyperlink{ref-harrison_influence_2020}{Harrison et al. 2020};
\protect\hyperlink{ref-mott_amphistomy_1991}{Mott and Michaelson 1991}).
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
(\protect\hyperlink{ref-sachs_developmental_1974}{Sachs 1974}); however,
the developmental controls of stomatal patterning are poorly understood
and likely more complex than random development along the leaf surface.
Croxdale (\protect\hyperlink{ref-croxdale_stomatal_2000}{2000}){]}
reviews three developmental theories which attempt to explain stomatal
patterning in angiosperms: inhibition, cell lineage, and cell cycle,
ultimately arguing for a cell cycle based control of stomatal
patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
(\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott
1990}; \protect\hyperlink{ref-manter_ci_2004}{Manter 2004};
\protect\hyperlink{ref-carriqui_diffusional_2015}{Carriquí et al.
2015}). Low light decreasing CO\(_2\) demand by limiting electron
transport rate, leading to relatively high internal CO\(_2\)
concentration (\(C_\text{i}\)) and low \(A_\text{max}\)
(\protect\hyperlink{ref-kaiser_metabolic_2016}{Kaiser et al. 2016}). In
contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis
(\protect\hyperlink{ref-farquhar_biochemical_1980}{Farquhar, Caemmerer,
and Berry 1980}; \protect\hyperlink{ref-lehmeier_cell_2017}{Lehmeier et
al. 2017}). In this study, we focus primarily on how stomatal patterning
affects diffusion, ignoring boundary layer and mesophyll resistances.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length
(\protect\hyperlink{ref-parkhurst_diffusion_1994}{Parkhurst 1994}). As
the diffusion rate of CO\(_2\) though liquid is approximately
\(10^4\times\) slower than CO\(_2\) diffusion through air, mesophyll
resistance is generally thought to be primarily limited by liquid
diffusion (\protect\hyperlink{ref-aalto_three-dimensional_2002}{Aalto
and Juurola 2002}; \protect\hyperlink{ref-evans_resistances_2009}{J. R.
Evans et al. 2009}), but diffusion through the IAS has also been shown
to be a rate limiting process because the tortuous, disjunct nature of
the IAS can greatly increase diffusion path lengths
(\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}). Additionally, tortuosity is higher
in horizontal directions (parallel to leaf surface) than vertical
directions (perpendicular to leaf surface) because of the cylindrical
shape and vertical arrangement of pallisade mesophyll cells
(\protect\hyperlink{ref-earles_beyond_2018}{Earles et al. 2018};
\protect\hyperlink{ref-harwood_understanding_2021}{Harwood,
Théroux‐Rancourt, and Barbour 2021}). However, the ratio of lateral to
vertical diffusion rate is still largely unknown
(\protect\hyperlink{ref-morison_lateral_2005}{Morison et al. 2005};
\protect\hyperlink{ref-pieruschka_lateral_2005}{R. Pieruschka 2005};
\protect\hyperlink{ref-pieruschka_lateral_2006}{Roland Pieruschka et al.
2006}). Depending on the thickness of the leaf, porosity of the leaf
mesophyll, tortuosity of the IAS, and lateral to vertical diffusion rate
ratio, minimizing diffusion path length for CO\(_2\) via optimally
distributed stomata may yield significant increases in CO\(_2\) supply
for photosynthesis and higher \(A_\text{max}\).

@@

\emph{Arabidopsis thaliana} plants were grown in three different light
environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Once leaves were mature, we
captured images of the abaxial and adaxial leaf surfaces using XXX. We
captured 132 images in total, making 66 abaxial-adaxial image pairs. We
measured stomatal position and size using ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

\noindent \(A_\text{leaf}\) is leaf area visible in the sampled field
and \(n_\text{stomata}\) the number of stomata. \(\overline{D}_E\) is
the theoretical average distance to the nearest neighbor of each stomate
if stomata were uniformly randomly distributed
(\protect\hyperlink{ref-clark_distance_1954}{Clark and Evans 1954}).
\(\overline{D}_O\) calculated for each synthetic data set is:

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1
(\protect\hyperlink{ref-evans_spatialeco_2023}{J. S. Evans and Murphy
2023}). The observed stomatal distribution is dispersed relative to a
uniform random distribution if the observed \(NNI\) is greater than 95\%
of the synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg
(\protect\hyperlink{ref-benjamini_controlling_1995}{\textbf{benjamini\_controlling\_1995?}})
false discovery rate procedure implemented in the \emph{R} package
\textbf{multtest} version 2.56.0
(\protect\hyperlink{ref-pollard_multiple_2005}{\textbf{pollard\_multiple\_2005?}}).

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian generalized non-linear multilevel model
fit with the \emph{R} package \textbf{brms} version 2.15.0
(\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017},
\protect\hyperlink{ref-burkner_advanced_2018}{2018}) and \emph{Stan}
version 2.27.0
(\protect\hyperlink{ref-stan_development_team_stan_2023}{\textbf{stan\_development\_team\_stan\_2023?}}).
Stomatal zone area was calculated using Voronoi tessellation
(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
\(S_\text{area}\), is the region of the leaf surface whose distance to
stomate, \(S\), is less than the distance to any other stomate, \(S\).
Stomatal length was measured in ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Correlation between paired abaxial and adaxial leaf surfaces. Dashed line indicates no correlation. Weak positive correlations are not significantly different from null simulations. No differences in abaxial-adaxial correlation were observed between light levels according to an analysis of variance ($\alpha$ = 0.05).}
\label{fig:dual-surface}
\end{figure}

Across all light treatments and leaf surfaces, stomatal length and
stomatal area were weakly positively correlated, indicating some support
for our hypothesis that stomata are selected for an optimal operational
aperture (\autoref{fig:length-area}).

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area. Lines of best fit computed using a bayesian generalized non-linear multilevel model. Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
\label{fig:length-area}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}

% Set document class options
\documentclass[webpdf,large,modern,unnumsec,namedate]{oup-authoring-template}

% one column
\onecolumn

%\usepackage{showframe}

% line numbers

@@

% From Pandoc template for its feature
\usepackage{xcolor}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous Arabidopsis thaliana Incurs No Photosynthetic Advantage},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

% Counters for addresses and footnotes
\newcounter{correspcnt} % For author footnotes
\renewcommand*{\thecorrespcnt}{\fnsymbol{correspcnt}}
\newcounter{addrcnt} % For author addresses

% Macros for dealing with affiliations, footnotes, etc.
\makeatletter

\def\MyNewLabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}

\def\MyRef#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}

@@

\newcommand*\addrlblbycode[1]{%
  \ifcounter{ADDRLBL@#1}
    {}
    {\refstepcounter{addrcnt}\newcounter{ADDRLBL@#1}\setcounter{ADDRLBL@#1}{\value{addrcnt}}}%
    \arabic{ADDRLBL@#1}%
}

\newcommand*\addrbycode[1]{%
  \ifcounter{ADDR@#1}
    {}
    {\newcounter{ADDR@#1}%
     \address[\addrlblbycode{#1}]{\MyRef{ADDRTXT}{#1}}}%
}

\newcommand*\corresplblbycode[1]{%
  \ifcounter{CORRESPLBL@#1}
    {}
    {\refstepcounter{correspcnt}\newcounter{CORRESPLBL@#1}\setcounter{CORRESPLBL@#1}{\value{correspcnt}}}%
    \fnsymbol{CORRESPLBL@#1}%
}

\newcommand*\correspbycode[1]{%
  \ifcounter{CORRESP@#1}
    {}
    {\newcounter{CORRESP@#1}%
     \corresp[\corresplblbycode{#1}]{\MyRef{CORRESPTXT}{#1}}}%
}

\makeatother

% Add missing \city command mentioned in documentation but absent from cls
\providecommand\city[1]{#1}

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{Colgate University}{%
  13 Oak Drive, Hamilton, NY 13346%
 }
   \MyNewLabel{ADDRTXT}{University of Hawaii at Manoa}{%
  2500 Campus Rd, Honolulu, HI 96822%
 }

% Create labels for Footnotes if they are given in Elsevier format
\MyNewLabel{CORRESPTXT}{1}{Equal contribution}
\MyNewLabel{CORRESPTXT}{2}{Current email address:
\href{mailto:cat@example.com}{cat@example.com}}

% Pandoc header-include feature
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
% Pandoc header-include feature
\usepackage{booktabs}

@@

\journaltitle{Journal Title Here}
\DOI{DOI HERE}
\copyrightyear{YYYY}
\pubyear{YYYY}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

\title[]{Spatial Coordination Of Stomatal Patterning Between Leaf
Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No
Photosynthetic Advantage}

\newcounter{thisauthcorresp} % For storage if author is corresponding author
\newcounter{thisauththanks} % For storage if author has thanks

\author[%
\addrlblbycode{Colgate University}%
,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
%
%
]{Jacob L. Watts}

\addrbycode{Colgate University}

\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:jwatts@colgate.edu}{\nolinkurl{jwatts@colgate.edu}}}

\author[%
\addrlblbycode{ETH}%
%
%
%
]{Graham Dow}

\addrbycode{ETH}

\author[%
\addrlblbycode{University of California Davis}%
%
%
,\corresplblbycode{1}%
]{Thomas N. Buckley}

\addrbycode{University of California Davis}

\correspbycode{1}

\author[%
\addrlblbycode{University of Hawaii at Manoa}%
%
%
,\corresplblbycode{1}%
]{Christopher D. Muir}

\addrbycode{University of Hawaii at Manoa}

\correspbycode{1}

% Add author mark
\authormark{Jacob L. Watts et al.}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

\abstract{
This is the abstract. It consists of two paragraphs.}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry points through which gas exchange occurs despite making up a small
percentage of the leaf area \citep{lange_responses_1971}. Stomata
consist of two guard cells which open and close upon changes in turgor
pressure or hormonal cues \citep{mcadam_linking_2016}. The stomatal pore
leads to an internal space known as the substomatal cavity where gases
contact the mesophyll. Once in the mesophyll, CO\(_2\) diffuses
throughout a network of intercellular air space (IAS) and into mesophyll
cells where CO\(_2\) assimilation (\(A\)) occurs within the chloroplasts
\citep{lee_diffusion_1964}. Stomatal conductance and transpiration are
determined by numerous environmental and anatomical parameters such as
vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf
water potential, IAS geometry, mesophyll cell anatomy, and stomatal
anatomy.

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
However, stomatal anatomy may be partially constrained by physical and
developments limits on phenotypic expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
\citep{geisler_oriented_2000, dow_physiological_2014}; however some
species (notably in \emph{Begonia}) appear to benefit from overlapping
vapor shells caused by stomatal clustering
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Stomatal traits also vary adaptively in different environments. Stomatal
density positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
\citep{jordan_environmental_2015}. Size positively co-varies with genome
size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
density \citep{camargo_density_2011}. Total stomatal area (size
\(\times\) density) is optimized for operational conductance
(\(gs_\text{s,op}\)) rather than maximum conductance
(\(gs_\text{s,max}\)) such that stomatal apertures are most responsive
to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(gs_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(gs_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. In most species, stomata occur on the
abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
high light environments with constant or intermittent access to
sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are poorly understood and likely more complex than
random development along the leaf surface.
\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
which attempt to explain stomatal patterning in angiosperms: inhibition,
cell lineage, and cell cycle, ultimately arguing for a cell cycle based
control of stomatal patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreasing CO\(_2\) demand by limiting electron transport
rate, leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
this study, we focus primarily on how stomatal patterning affects
diffusion, ignoring boundary layer and mesophyll resistances.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length \citep{parkhurst_diffusion_1994}. As the diffusion
rate of CO\(_2\) though liquid is approximately \(10^4\times\) slower
than CO\(_2\) diffusion through air, mesophyll resistance is generally
thought to be primarily limited by liquid diffusion
\citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
diffusion through the IAS has also been shown to be a rate limiting
process because the tortuous, disjunct nature of the IAS can greatly
increase diffusion path lengths \citep{harwood_understanding_2021}.
Additionally, tortuosity is higher in horizontal directions (parallel to
leaf surface) than vertical directions (perpendicular to leaf surface)
because of the cylindrical shape and vertical arrangement of pallisade
mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
However, the ratio of lateral to vertical diffusion rate is still
largely unknown
\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006}.
Depending on the thickness of the leaf, porosity of the leaf mesophyll,
tortuosity of the IAS, and lateral to vertical diffusion rate ratio,
minimizing diffusion path length for CO\(_2\) via optimally distributed
stomata may yield significant increases in CO\(_2\) supply for
photosynthesis and higher \(A_\text{max}\).

@@

\emph{Arabidopsis thaliana} plants were grown in three different light
environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Once leaves were mature, we
captured images of the abaxial and adaxial leaf surfaces using XXX. We
captured 132 images in total, making 66 abaxial-adaxial image pairs. We
measured stomatal position and size using ImageJ
\citep{schneider_nih_2012}.

@@

\noindent \(A_\text{leaf}\) is leaf area visible in the sampled field
and \(n_\text{stomata}\) the number of stomata. \(\overline{D}_E\) is
the theoretical average distance to the nearest neighbor of each stomate
if stomata were uniformly randomly distributed
\citep{clark_distance_1954}. \(\overline{D}_O\) calculated for each
synthetic data set is:

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is dispersed relative to a uniform
random distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.56.0 \citep{wong_multiple_2005}.

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian linear mixed-effects model fit with the
\emph{R} package \textbf{brms} version 2.20.4
\citep{burkner_brms_2017, burkner_advanced_2018} and \emph{Stan} version
2.33.1 \citep{stan_development_team_stan_2023}. Stomatal zone area was
calculated using Voronoi tessellation
(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
\(S_\text{area}\), is the region of the leaf surface whose distance to
stomate, \(S\), is less than the distance to any other stomate, \(S\).
Stomatal length was measured in ImageJ \citep{schneider_nih_2012}. We
modeled fixed effects of surface, light treatment, stomatal length, and
their 2- and 3-way interactions on \(\sqrt{S_\text{area}}\). We included
random intercepts, random effects of surface, random slopes, and random
surface-by-slope interactions within both plant and individual to
account for nonindependence of stomata within the same plant or
individual. We also modeled residual variance as a function of light
treatment. We sampled the posterior distribution from 4 chains with 1000
iterations each after 1000 warmup iterations. We calculated convergence
diagnostics (\(\hat{R}\)) and effective sample sizes following
\citet{vehtari_rank-normalization_2021}. We estimated the marginal slope
and 95\% highest posterior density (HPD) intervals between stomatal
length and \(\sqrt{S_\text{area}}\) using the \emph{emtrends} function
in the \emph{R} package \textbf{emmeans} version 1.8.8
\citep{lenth_emmeans_2023}.

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces. Dashed line indicates zero correlation. Weak positive correlations are not significantly different from zero after correcting for multiple comparisons. The correlation does not differ among light treatments.}
\label{fig:dual-surface}
\end{figure}

\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
\subsection{Larger stomata supply larger mesophyll
volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}
